Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-02 appendix.tex
fix deprecated $$; tex kinda sux :(

We first fix notation.
\begin{definition}
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$.
Let the mapping from site patterns to site splits
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
$$
be
$$
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \alignmentColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\alignmentColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet)
    \end{array}
\right.
$$
and the mapping from ancestral states and tip states to ancestral state splits
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
be
$$
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \ancestralStateColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\ancestralStateColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet).
    \end{array}
\right.
$$
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.

For the $i$th factor of \eqref{eq:full_likelihood},
$$
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ depending on $i$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.

@@

Let
$$
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
$$
\theta(e) = 1-2p(e)
$$
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

@@

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
$$
\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
$$
\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

@@

\begin{proof}
In general,
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
$$
is an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
$$
Similarly,
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
$$
is a lower bound.

@@

$$

@@

$$

>>>>

We first fix notation.
\begin{definition}
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$.
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \alignmentColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\alignmentColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet)
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \ancestralStateColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\ancestralStateColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet).
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
\]
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.

For the $i$th factor of \eqref{eq:full_likelihood},
\[
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
\]
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ depending on $i$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.

@@

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
\]
and the partial log-likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
\]
so that \eqref{eq:site_pattern_profile_likelihood_mean} is
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

@@

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
\]

@@

\begin{proof}
In general,
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
\]
is an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
\]
and
\[
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\]
Similarly,
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
\]
is a lower bound.

@@

\[

@@

\]
